Baseline Drift Estimation for Air Quality Data Using Quantile Trend Filtering

### 2019-04-09 Manuscript/detrendify_aoas.tex
responding to Joe's comments, updated npqw for Fig 8 and 9

\section{Introduction}
\label{sec:intro}

@@

In the last decade, low cost and portable air quality sensors have enjoyed dramatically increased usage. These sensors can provide an un-calibrated measure of a variety of pollutants in near real time, but deriving meaningful information from sensor data remains a challenge \citep{snyder2013changing}. The ``SPod" is a low-cost sensor currently being investigated by researchers at the U.S. Environmental Protection Agency to detect volatile organic compound (VOC) emissions from industrial facilities \citep{thoma2016south}. Due to changes in temperature and relative humidity the output signal exhibits a slowly varying baseline drift on the order of minutes to hours. Figure \ref{fig:raw_spod} provides an example of measurements from three SPod sensors co-located at the border of an industrial facility. All of the sensors respond to the pollutant signal, which is illustrated by the three sharp transient spikes at 11:32, 14:10, and 16:03. However, the baseline drift varies from one sensor to another, obscuring the detection of the peaks that alert the intrusion of pollutants. We show later that by estimating the baseline drift in each sensor %using $\ell_1$-quantile-trend filtering
and  removing it from the observed signals, peaks can be reliably detected from concordant residual signals from a collection of SPods using a simple data-driven thresholding strategy. Thus, accurately demixing a noisy observed vector $\V{y}$ into a slowly varying component $\V{\theta}$ and a transient component $\V{s}$ can lead to greatly improved and simplified downstream analysis.

@@

We combine the ideas of quantile regression and trend filtering. For a single quantile level $\tau$, the quantile trend filtering problem is given in \Eqn{quantile_trend}. As with classic quantile regression, the quantile trend filtering problem is a linear program which can be solved by a number of methods. We want to estimate multiple quantiles simultaneously and to ensure that our quantile estimates are valid by enforcing the constraint that if $\tau_2 > \tau_1$ then $Q(\tau_2) \ge Q(\tau_1)$ where $Q$ is the quantile function of $\V{y}$. Even if a single quantile is ultimately desired, ensuring non-crossing allows information from nearby quantiles to be used to improve the estimates. Given quantiles $\tau_1 < \tau_2 < \ldots < \tau_J$, the optimization problem becomes
\begin{eqnarray*}
	\label{eq:noncross_trend}
	&&\underset{\M{\Theta} \in \mathcal{C}}{\text{minimize}}\; \sum_{j=1}^J \left [\rho_{\tau_j}(\V{y} - \V{\theta}_{j}) +
	\lambda_j \lVert \Mn{D}{k+1} \V{\theta}_j \rVert_1 \right ] \\
\end{eqnarray*}
where $\M{\Theta} = \begin{pmatrix} \V{\theta}_1 & \V{\theta}_2 & \cdots & \V{\theta}_J \end{pmatrix} \in \Real^{n \times J}$ is a matrix whose $j$th column corresponds to the $j$th quantile signal $\V{\theta}_j$ and the set $\mathcal{C} = \{\M{\Theta} \in \Real^{n \times J}: \ME{\theta}{ij} \leq \ME{\theta}{ij'} \;\text{for $j \leq j'$}\}$ encodes the non-crossing quantile constraints. The additional constraints are linear in the parameters, so the non-crossing quantile trends can still be estimated by a number of available solvers. We allow for the possibilty that the degree of smoothness in the trends varies by quantile and thus allow the smoothing parameter as well. In the rest of this paper, we report numerical results using the commercial solver Gurobi \citep{gurobi} and its R package implementation. However, we could easily substitute a free solver such as the Rglpk package by \cite{rglpk}.

@@

 As the size of the data increases, all methods require additional computation time and at some point the penalty matrices will no longer fit into memory. In our application to air quality sensor data, measurements are recorded every second resulting in 86,400 observations per day. This number of observations is already too large to use with currently available R packages used for estimating quantile trends \citep{fields, quantreg}. To our knowledge, no one has addressed the problem of finding smooth quantile trends of series that are too large to be processed simultaneously. We propose a divide-and-conquer approach via an ADMM algorithm for solving large problems in a piecewise fashion.

@@

Returning to our constrained windows problem giving in \Eqn{quantile_windows}, let $\Mn{\Omega}{w}$ denote the Lagrange multiplier matrix for the $w$th consensus constraint, namely $\Mn{\theta}{w} = \Mn{U}{w}\M{\theta}$, and let $\Vn{\omega}{w}_{j}$ denote its $j$th column.

@@

\label{sec:application}
The low-cost ``SPod" air quality sensors output a time series that includes a slowly varying baseline,
the sensor response to pollutants, and high frequency random noise. These sensors record measurements every second and are used to monitor pollutant concentrations at the fence lines of industrial facilities. Time points with high concentrations are identified and compared with concurrent wind direction and speed. Ideally, three co-located and time aligned sensors (as shown in \Fig{raw_spod}) responding to a pollutant plume would result in the same signal classification after baseline trend removal and proper threshold choice. We first illustrate the difference between our \texttt{detrend\_eBIC} method,  hereafter referred to as \texttt{detrendr}, and \texttt{qsreg} using a small subset of measurements from \Fig{raw_spod} (\Sec{short-app}). We then compare the methods on the complete day shown in \Fig{raw_spod}, estimating trends by applying the \texttt{qsreg} method to 2 hour non-overlapping windows of the data and \Alg{admm} to the entire day. We focus on this day because data from three sensors was available. Finally, we examine an entire week of measurements from two co-located sensors (\Sec{long-app}).

@@

We compare our \texttt{detrendr} method with the \texttt{qsreg} method on a subset of the SPod data (n=7200) both to facilitate visualization and because the \texttt{qsreg} method cannot handle all 24 hours simultaneously. We estimate the baseline trend using $\tau = \{0.01, 0.05, 0.1\}$ and compare three thresholds for classifying the signal. The thresholds are calculated as the 90\textsuperscript{th}, 95\textsuperscript{th}, and 99\textsuperscript{th} quantiles of the de-trended series for each SPod. If there is signal present in the dataset, values above these thresholds should occur simultaneously on all three SPods. We do not use class-averaged accuracy to compare the signal classifications because we do not have a reference value to define as the ``true" signal. Instead, we compute the variation of information (VI) which compares the similarity between the two classifications. Given the signal classifications for SPods a and b, $\delta^{(a)}_i \in \{0,1\}$ and $\delta^{(b)}_i\in\{0,1\}$, the VI is defined as:
\begin{eqnarray*}
	r_{ij} & = & \frac{1}{n}\sum_t \One\left(\delta^{(a)}_i = i  \cap \delta^{(b)}_i = j\right)\\
	\VI(s_a, s_b) & = & -\sum_{i,j} r_{ij} \left[ \log \left(\frac{r_{ij}}{\frac{1}{n}\sum_t \One(\delta^{(a)}_i = i)}\right) +
	\log \left(\frac{r_{ij}}{\frac{1}{n}\sum_t \One(\delta^{(b)}_i = j)}\right) \right]
\end{eqnarray*}
The $\VI$ is a distance metric for measuring similarity of classifications and will be 0 if the classifications are identical and increase as the classifications become more different.

@@

\subsection{Long series of air quality measurements}
\label{sec:long-app}
\Alg{admm} was used to remove the baseline drift from the full day of data  (\Fig{raw_spod}) consisting of 86,401 observations per SPod and compared to the series detrended using the \texttt{qsreg} trends estimated using non-overlapping 2 hour windows. As in the shorter illustration, the \texttt{detrendr} method results in generally higher VI scores than the \texttt{qsreg} method (\Fig{vi_day}). The \texttt{detrendr} method also results in better correlation in the detrended series as is illustrated in (\Fig{scatter_day}). The Spearman correlation coefficients for SPods a and b, SPods a and c, and SPods b and c after removing the 5\textsuperscript{th} quantile trend using \texttt{detrendr} were 0.37, 0.75, and 0.43, compared with 0.07, 0.24, and 0.16 using \texttt{qsreg}. The noise variance was higher for SPod b than for SPods a and c resulting in lower correlation and higher VI values for the ab and bc metrics compared with ac.

@@

Finally we estimated the quantile trends for 7 days of measurements of two co-located SPods. \Fig{vi_week} demonstrates the improvement in classification similarity when using \texttt{detrendr}. Each point represents a day of measurements and all points that fall below the dashed line have more similar classifications using \texttt{detrendr} compared to \texttt{qsreg}.

\begin{figure}
	\includegraphics[width = 0.7\linewidth]{Figures/VI_by_day.png}
	\caption{Variation of Information (VI) for \texttt{detrendr} and \texttt{qsreg} by quantile trend and threshold. Each point represents a full day of data. The dashed line represents y=x. In most cases \texttt{detrendr} results in a lower VI than \texttt{qsreg}.}
	\label{fig:vi_week}
\end{figure}

@@

We have expanded the quantile trend filtering method by implementing a non-crossing constraint and a new algorithm for processing large series, and proposing a modified criteria for smoothing parameter selection. Furthermore we have demonstrated the utility of quantile trend filtering in both simulations and applied settings. Our ADMM algorithm for large series both reduces the computing time and allows trends to be estimated on series that cannot be estimated simultaneously while our scaled extended BIC criteria was shown to provided better estimated of quantile trends in series with and without a signal component. We have also shown that the baseline drift in low cost air quality sensors can be removed through estimating quantile trends.

@@

\label{sec:conc}

>>>>

\section{Introduction}
\label{sec:intro}
In the last decade, low cost and portable air quality sensors have enjoyed dramatically increased usage. These sensors can provide an un-calibrated measure of a variety of pollutants in near real time, but deriving meaningful information from sensor data remains a challenge \citep{snyder2013changing}. The ``SPod" is a low-cost sensor currently being investigated by researchers at the U.S. Environmental Protection Agency to detect volatile organic compound (VOC) emissions from industrial facilities \citep{thoma2016south}. Due to changes in temperature and relative humidity the output signal exhibits a slowly varying baseline drift on the order of minutes to hours. Figure \ref{fig:raw_spod} provides an example of measurements from three SPod sensors co-located at the border of an industrial facility. All of the sensors respond to the pollutant signal, which is illustrated by the three sharp transient spikes at 11:32, 14:10, and 16:03. However, the baseline drift varies from one sensor to another, obscuring the detection of the peaks that alert the intrusion of pollutants. We show later that by estimating the baseline drift in each sensor and  removing it from the observed signals, peaks can be reliably detected from concordant residual signals from a collection of SPods using a simple data-driven thresholding strategy. Thus, accurately demixing a noisy observed vector into a slowly varying component and a transient component can lead to greatly improved and simplified downstream analysis.

@@

@@

We combine the ideas of quantile regression and trend filtering. For a single quantile level $\tau$, the quantile trend filtering problem is given in \Eqn{quantile_trend}. As with classic quantile regression, the quantile trend filtering problem is a linear program which can be solved by a number of methods. We want to estimate multiple quantiles simultaneously and to ensure that our quantile estimates are valid by enforcing the constraint that if $\tau_2 > \tau_1$ then $Q(\tau_2) \ge Q(\tau_1)$ where $Q$ is the quantile function of $\V{y}$. Even if a single quantile is ultimately desired, ensuring non-crossing allows information from nearby quantiles to be used to improve the estimates. Given quantiles $\tau_1 < \tau_2 < \ldots < \tau_J$, the optimization problem becomes
\begin{eqnarray*}
	\label{eq:noncross_trend}
	&&\underset{\M{\Theta} \in \mathcal{C}}{\text{minimize}}\; \sum_{j=1}^J \left [\rho_{\tau_j}(\V{y} - \V{\theta}_{j}) +
	\lambda_j \lVert \Mn{D}{k+1} \V{\theta}_j \rVert_1 \right ] \\
\end{eqnarray*}
where $\M{\Theta} = \begin{pmatrix} \V{\theta}_1 & \V{\theta}_2 & \cdots & \V{\theta}_J \end{pmatrix} \in \Real^{n \times J}$ is a matrix whose $j$th column corresponds to the $j$th quantile signal $\V{\theta}_j$ and the set $\mathcal{C} = \{\M{\Theta} \in \Real^{n \times J}: \ME{\theta}{ij} \leq \ME{\theta}{ij'} \;\text{for $j \leq j'$}\}$ encodes the non-crossing quantile constraints. The additional constraints are linear in the parameters, so the non-crossing quantile trends can still be estimated by a number of available solvers. We allow for the possibility that the degree of smoothness in the trends varies by quantile by allowing the smoothing parameter to vary with quantile as well. In the rest of this paper, we report numerical results using the commercial solver Gurobi \citep{gurobi} and its R package implementation. However, we could easily substitute a free solver such as the Rglpk package by \cite{rglpk}.

@@

 As the size of the data increases, computation time becomes prohibitive \Fig{timing}. In our application to air quality sensor data, measurements are recorded every second resulting in 86,400 observations per day. This number of observations is already too large to use with currently available R packages used for estimating quantile trends \citep{fields, quantreg}. To our knowledge, no one has addressed the problem of finding smooth quantile trends of series that are too large to be processed simultaneously. We propose a divide-and-conquer approach via an ADMM algorithm for solving large problems in a piecewise fashion.

@@

Returning to our constrained windows problem giving in \Eqn{quantile_windows}, let $\Mn{\Omega}{w}$ denote the Lagrange multiplier matrix for the $w$th consensus constraint, namely $\Mn{\Theta}{w} = \Mn{U}{w}\M{\Theta}$, and let $\Vn{\omega}{w}_{j}$ denote its $j$th column.

@@

\label{sec:application}
The low-cost ``SPod" air quality sensors output a time series that includes a slowly varying baseline,
the sensor response to pollutants, and high frequency random noise. These sensors record measurements every second and are used to monitor pollutant concentrations at the perimeter of industrial facilities. Time points with high concentrations are identified and compared with concurrent wind direction and speed. Ideally, three co-located and time aligned sensors (as shown in \Fig{raw_spod}) responding to a pollutant plume would result in the same signal classification after baseline trend removal and proper threshold choice. We first illustrate the difference between our \texttt{detrend\_eBIC} method,  hereafter referred to as \texttt{detrendr}, and \texttt{qsreg} using data from 13:10 to 15:10 from \Fig{raw_spod} (\Sec{short-app}). We then compare the methods on the complete day shown in \Fig{raw_spod}, estimating trends by applying the \texttt{qsreg} method to 2 hour non-overlapping windows of the data and \Alg{admm} to the entire day. We focus on this day because data from three sensors was available. Finally, we examine an entire week of measurements from two co-located sensors (\Sec{long-app}).

@@

We compare our \texttt{detrendr} method with the \texttt{qsreg} method on a two-hour subset of one-second  SPod data (n=7200) both to facilitate visualization and because the \texttt{qsreg} method cannot handle all 24 hours simultaneously. We estimate the baseline trend using $\tau = \{0.01, 0.05, 0.1\}$ and compare three thresholds for classifying the signal. The thresholds are calculated as the 90\textsuperscript{th}, 95\textsuperscript{th}, and 99\textsuperscript{th} quantiles of the de-trended series for each SPod. If there is signal present in the dataset, values above these thresholds should occur simultaneously on all three SPods. We do not use class-averaged accuracy to compare the signal classifications because we do not have a reference value to define as the ``true" signal. Instead, we compute the variation of information (VI) which compares the similarity between the two classifications. Given the signal classifications for SPods a and b, $\delta^{(a)}_i \in \{0,1\}$ and $\delta^{(b)}_i\in\{0,1\}$, the VI is defined as:
\begin{eqnarray*}
	r_{ij} & = & \frac{1}{n}\sum_t \One\left(\delta^{(a)}_i = i  \cap \delta^{(b)}_i = j\right)\\
	\VI(s_a, s_b) & = & -\sum_{i,j} r_{ij} \left[ \log \left(\frac{r_{ij}}{\frac{1}{n}\sum_t \One(\delta^{(a)}_i = i)}\right) +
	\log \left(\frac{r_{ij}}{\frac{1}{n}\sum_t \One(\delta^{(b)}_i = j)}\right) \right]
\end{eqnarray*}
The $\VI$ is a distance metric for measuring similarity of classifications and will be 0 if the classifications are identical and increase as the classifications become more different.

@@

\subsection{Long series of air quality measurements}
\label{sec:long-app}
\Alg{admm} was used to remove the baseline drift from the full day of data  (\Fig{raw_spod}) consisting of 86,401 observations per SPod and compared to the series detrended using the \texttt{qsreg} trends estimated using non-overlapping 2 hour windows. As in the shorter illustration, the \texttt{detrendr} method results in generally lower VI scores than the \texttt{qsreg} method (\Fig{vi_day}). The \texttt{detrendr} method also results in better correlation in the detrended series as is illustrated in (\Fig{scatter_day}). The Spearman correlation coefficients for SPods a and b, SPods a and c, and SPods b and c after removing the 5\textsuperscript{th} quantile trend using \texttt{detrendr} were 0.37, 0.75, and 0.43, compared with 0.07, 0.24, and 0.16 using \texttt{qsreg}. The noise variance was higher for SPod b than for SPods a and c resulting in lower correlation and higher VI values for the ab and bc metrics compared with ac.

@@

Finally we estimated the quantile trends for 7 days of measurements of two co-located SPods. \Fig{vi_week} demonstrates the improvement in classification similarity when using \texttt{detrendr}. Each point represents a day of measurements and all points that fall below the dashed line have more similar classifications using \texttt{detrendr} compared to \texttt{qsreg}. The improvement of \texttt{detrendr} over \texttt{qsreg} is more severe at lower thresholds. This indicates that \texttt{detrendr} gives greater agreement on signal classification when the methods are tuned to deliver positive classifications more frequently.

\begin{figure}
	\includegraphics[width = 0.6\linewidth]{Figures/VI_by_day.png}
	\caption{Variation of Information (VI) for \texttt{detrendr} and \texttt{qsreg} by quantile trend and threshold. Each point represents a full day of data. The dashed line represents y=x. In most cases \texttt{detrendr} results in a lower VI than \texttt{qsreg}.}
	\label{fig:vi_week}
\end{figure}

@@

We have expanded the quantile trend filtering method by implementing a non-crossing constraint and a new algorithm for processing large series, and proposing a modified criteria for smoothing parameter selection. Furthermore we have demonstrated the utility of quantile trend filtering in both simulations and applied settings. Our ADMM algorithm for large series both reduces the computing time and allows trends to be estimated on series that cannot be estimated simultaneously while our scaled extended BIC criteria was shown to provided better estimated of quantile trends in series with and without a signal component.

In our application to low cost air quality sensor data, we have shown that the baseline drift in low cost air quality sensors can be removed through estimating quantile trends, but the data size was too large for existing methods to be computationally feasible. While \texttt{qsreg} cannot feasibly handle more than a few hour windows of data, our new methods were able to process 24 hours simultaneously and deliver signal classifications that were more consistent between the two sensors for a week of data (168 hours).

@@

\section{Acknowledgments}
The authors greatly appreciate Eben Thoma at the US Environmental Protection Agency for providing the SPod datasets. The authors are also thankful for the Oak Ridge Institute of Science Fellowship that partially supported this work. Plots were created using ggplot2 \citep{ggplot2}.
